---
title: "Final Project"
author: '41586838'
date: "2022-11-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

List of Things to do:
1. use strings to rename years done
2. take 25 richest countries compare to everybody else
3. create an indicator variable
    variable called rich country
    give all rich countries 1 and everyone else a 0
4. take CO2 emisions/people
5. plot 2 lines
6. how does that look over time for rich countries


second: take C02 per income

regression table
regress CO2 value on income per person
then add fixed effect for year and country
control for variation across time and country
detrend a lot of the data



```{r}
library(tidyverse)
library(tidyverse)
library(ggplot2)
library(gt)
library(gtExtras)
library(lubridate)
library(scales)
setwd("/Users/nsusser/Dropbox/PSCI1800")
WBdata <- read_csv("Data/Raw/World Bank FP.csv") 
WBdata %>%
  head(10)
View(WBdata)
```

```{r}
clean_data <- WBdata %>%
  pivot_longer(c(`2000 [YR2000]`:`2020 [YR2020]`),
                                 names_to = "year",
                                 values_to = "values") %>%
  filter(!(values==".."))%>%
  mutate(series = ifelse(`Series Code` == "EN.ATM.CO2E.KT","CO2", ifelse(`Series Code` == "SP.POP.TOTL","POP","NNI"))) %>%
  mutate(num_values = as.numeric(values))
  
  
for (i in 1:length(clean_data$`Country Name`)) {
  clean_data$year[i] <- str_sub(clean_data$year[i], 1, 4)
}
  

View(clean_data)

summary(clean_data$num_values)
```

```{r}
CO2_data <- clean_data %>%
  filter(series == "CO2") %>%
  select(`Country Name`, `Country Code`, year, num_values) %>%
  rename(CO2_values = num_values)
#View(CO2_data)

NNI_data <- clean_data %>%
  filter(series == "NNI") %>%
  select(`Country Name`, `Country Code`, year, num_values) %>%
  rename(NNI_values = num_values)
#View(NNI_data)

POP_data <- clean_data %>%
  filter(series == "POP") %>%
  select(`Country Name`, `Country Code`, year, num_values) %>%
  rename(POP_values = num_values)
#View(POP_data)

combined_data_1 <- inner_join(x = NNI_data, y = POP_data, by = c("Country Name","Country Code", "year"))

combined_data <- inner_join(x = combined_data_1, y = CO2_data, by = c("Country Name","Country Code", "year"))
View(combined_data)
```

```{r}
income_per_capita <- combined_data %>%
  mutate(ipc = NNI_values / POP_values) %>%
  group_by(`Country Code`) %>%
  summarise(avg_inc = mean(ipc)) %>%
  arrange(desc(avg_inc))

rich_countries <- income_per_capita %>%
  filter(avg_inc > 30000)
View(rich_countries)

mid_countries <- income_per_capita %>%
  filter(avg_inc > 3112 & avg_inc < 30000)
mid_countries

wealth_label <- combined_data %>%
  filter(`Country Code` != "MMR") %>% # filter out myranmar
  mutate(ipc = NNI_values / POP_values) %>%
  mutate(cpc = CO2_values / POP_values) %>% # kt per person
  mutate(rich = ifelse(`Country Code` %in% mid_countries$`Country Code`,"M", ifelse(`Country Code` %in% rich_countries$`Country Code`, "R","P"))) %>%
  mutate(co2perdollar = (CO2_values *1000) / NNI_values) %>% #tons of CO2 /dollar
  mutate(dollarsperton =NNI_values/(CO2_values *1000)) %>% #tons of CO2 /dollar
  mutate(cpc_ipc = (CO2_values *1000) / ipc, 
         year = as.Date(year, format = "%Y"),
         year = year(year)) # tons of CO2 
View(wealth_label)
```

```{r}

wealth_label %>%
  group_by(year, rich) %>%
  summarise(avg_cpc = 1000*mean(cpc))


wealth_label %>%
  group_by(year, rich) %>%
  summarise(avg_cpc = 1000*mean(cpc)) %>% # average co2 by rich and poor countries tons
  ggplot(aes(y = avg_cpc, x = year, color = as.factor(rich))) +
  geom_point() +
  ylab("CO2(tons) Per Capita") + 
  xlab("Year") +
  labs(color = "Country Wealth") +
  scale_color_discrete(labels=c('Middle','Poor', 'Rich')) +
  ggtitle("CO2 Per Capita Over Time")+ 
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme_bw()

wealth_label %>%
  group_by(year, rich) %>%
  summarise(avg_ipc = mean(ipc))

wealth_label %>%
  group_by(year, rich) %>%
  summarise(avg_ipc = mean(ipc)) %>% # average co2 by rich and poor countries tons
  ggplot(aes(y = avg_ipc, x = year, color = as.factor(rich))) +
  geom_line() +
  ylab("Income Per Capita") + 
  xlab("Year") +
  labs(color = "Country Wealth") +
  scale_color_discrete(labels=c('Middle','Poor', 'Rich')) +
  ggtitle("Income Per Capita Over Time")+ 
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme_bw()
```

```{r}
tots <- wealth_label %>%
  group_by(`rich`,`Country Name`) %>%
  summarise(avg_cpc = 1000*mean(cpc, na.rm = T),
            avg_ipc = mean(ipc, na.rm = T))
summary(lm(avg_cpc ~ as.factor(rich), data = tots))
```
```{r}
tots <- wealth_label %>%
  group_by(`rich`,`Country Name`) %>%
  summarise(avg_cpc = 1000*mean(cpc, na.rm = T),
            avg_ipc = mean(ipc, na.rm = T))
summary(lm(avg_ipc ~ as.factor(rich), data = tots))
```

```{r}
w2000 <- wealth_label %>%
  filter(year == 2000) %>%
  group_by(year,rich, `Country Name`) %>%
  summarise(avg_cpc = 1000*mean(cpc, na.rm = T),
            avg_ipc = mean(ipc, na.rm = T))
summary(lm(avg_cpc ~ as.factor(rich), data = w2000))
```

```{r}
w2019 <- wealth_label %>%
  filter(year == 2019) %>%
  group_by(year,rich,`Country Name`) %>%
  summarise(avg_cpc = 1000*mean(cpc, na.rm = T),
            avg_ipc = mean(ipc, na.rm = T))
summary(lm(avg_cpc ~ as.factor(rich), data = w2019))
```

```{r}
w2018 <- wealth_label %>%
  filter(year == 2018) %>%
  group_by(year,rich,`Country Name`) %>%
  summarise(avg_cpc = 1000*mean(cpc, na.rm = T),
            avg_ipc = mean(ipc, na.rm = T))
summary(lm(avg_cpc ~ as.factor(rich), data = w2018))
```


```{r}
wealth_label %>%
  filter(`Country Code` == c("USA", "PAK")) %>%
  ggplot(aes(y = co2perdollar, x = year, color = `Country Code`)) +
  geom_point()
```

```{r}
wealth_label %>%
  filter(`Country Code` == c("USA", "PAK")) %>%
  ggplot(aes(y = co2perdollar, x = year, group = `rich`, color = `rich`)) +
  geom_point()
```



```{r}
# two different calls 
  ggplot(data = wealth_label) +
  geom_point(aes(y = cpc, x = ipc, group = `rich`, color = `rich`)) +
  geom_smooth(aes(y = cpc, x = ipc), method = lm) +
  facet_wrap(~ year)
```

```{r}
wealth_label %>%
  ggplot(aes(y = ipc, x = cpc, group = `rich`, color = `rich`)) +
  geom_point() +
  facet_wrap(~ year)
```

```{r}
wealth_label %>%
  group_by(rich, year) %>%
  summarise(cpd = mean(co2perdollar, na.rm = T)) %>%
  ungroup() %>%
  ggplot() +
  geom_col(aes(y = cpd, x = year, fill = as.factor(`rich`)), position = "dodge2")
```

```{r}
wealth_label %>%
  group_by(rich, year) %>%
  summarise(cpd = mean(cpc_ipc, na.rm = T)) %>%
  ungroup() %>%
  ggplot() +
  geom_col(aes(y = cpd, x = year, fill = as.factor(`rich`)), position = "dodge2")
```

```{r}
wealth_label %>%
  group_by(rich, year) %>%
  summarise(cpd = mean(cpc_ipc, na.rm = T)) %>%
  ungroup() %>%
  ggplot() +
  geom_line(aes(y = cpd, x = year, color = as.factor(`rich`))) +
   geom_col(aes(y = cpd, x = year, fill = as.factor(`rich`)), position = "dodge2")
```



```{r}
wealth_label %>%
  filter(year == 2000 | year == 2004 | year == 2008 | year == 2012 | year == 2016 | year == 2019) %>%
  group_by(year,rich,`Country Name`) %>%
  summarise(avg_cpc = 1000*mean(cpc, na.rm = T),
            avg_ipc = mean(ipc, na.rm = T)) %>% # average co2 and income by rich and poor countries
  ggplot(aes(y = avg_cpc, x = avg_ipc)) +
  geom_point(aes(color = as.factor(rich))) +
  #geom_text(aes(label = `Country Name`)) +
  geom_smooth(method=lm) +
  facet_wrap(~ year) + 
  ylab("CO2(tons) per Capita") + 
  xlab("Net National Income Per Capita") +
  labs(color = "Country Wealth") +
  scale_color_discrete(labels=c('Poor', 'Rich')) +
  ggtitle("Higher Income Countries Produce More CO2")+ 
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme_bw()
```

```{r}
w2000 <- wealth_label %>%
  filter(year == 2000) %>%
  group_by(year,rich,`Country Name`) %>%
  summarise(avg_cpc = 1000*mean(cpc, na.rm = T),
            avg_ipc = mean(ipc, na.rm = T))
summary(lm(avg_cpc ~ avg_ipc, data = w2000))

```

```{r}
w2004 <- wealth_label %>%
  filter(year == 2004) %>%
  group_by(year,rich,`Country Name`) %>%
  summarise(avg_cpc = 1000*mean(cpc, na.rm = T),
            avg_ipc = mean(ipc, na.rm = T))
summary(lm(avg_cpc ~ avg_ipc, data = w2004))
```

```{r}
w2019 <- wealth_label %>%
  filter(year == 2019) %>%
  group_by(year,rich,`Country Name`) %>%
  summarise(avg_cpc = 1000*mean(cpc, na.rm = T),
            avg_ipc = mean(ipc, na.rm = T))
summary(lm(avg_cpc ~ avg_ipc, data = w2019))
```



```{r}
wealth_label %>%
  filter(`Country Name`%in% c("United States", "China", "India", "Pakistan", "United Kingdom", "Saudi Arabia", "Phillipines", "Bangladesh", "Haiti", "Ecuador")) %>%
  ggplot(aes(y = cpc * 1000, x = year)) +
  geom_point(aes(color = ipc)) #+
  #geom_text(aes(label = `Country Code`), position = position_dodge(0.1))
```

```{r}
View(wealth_label)
wealth_label %>%
  filter(`Country Code` == c("USA", "CHN", "IND")) %>%
  ggplot(aes(y = cpc * 1000, x = year, group = `Country Code`, color = ipc)) +
  geom_smooth()
```

```{r}
wealth_label %>%
  filter(`Country Name`%in% c("United States", "China", "India", "Pakistan", "United Kingdom", "Saudi Arabia", "Phillipines", "Bangladesh", "Haiti", "Ecuador")) %>%
  ggplot(aes(y = cpc * 1000, x = year)) +
  geom_line(aes(group = `Country Code`, color = ipc)) #+
  #geom_text(aes(label = `Country Code`), position = position_dodge(0.1))
```

```{r}
wealth_label %>%
  filter(`Country Name`%in% c( "China", "India", "Pakistan", "Phillipines", "Bangladesh", "Haiti", "Ecuador")) %>%
  ggplot(aes(y = cpc * 1000, x = year)) +
  geom_line(aes(group = `Country Code`, color = ipc))
```

```{r}
wealth_label %>%
  filter(rich == T) %>%
  ggplot(aes(y = cpc * 1000, x = year)) +
  geom_line(aes(group = `Country Code`, color = ipc)) +
  
  geom_text(aes(label = `Country Code`), position = position_dodge(0.1))
```



```{r}
wealth_label %>%
  group_by(year, rich) %>%
  summarise(avg_cpc = 1000*mean(cpc)) %>% # average co2 by rich and poor countries
  ggplot(aes(y = avg_cpc, x = year, color = as.factor(rich))) +
  geom_line() +
  ylab("CO2(tons) Per Capita") + 
  xlab("Year") +
  labs(color = "Country Wealth") +
  scale_color_discrete(labels=c('Poor', 'Rich')) +
  ggtitle("CO2 Per Capita Over Time")+ 
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme_bw()
```

```{r}
wealth_label %>%
  group_by(`Country Name`) %>%
  summarise(avg_co2 = 1000*mean(CO2_values)) %>%
  arrange(avg_co2)
```

```{r}
wealth_label %>%
  group_by(`Country Name`) %>%
  summarise(avg_inc = mean(NNI_values)) %>%
  arrange(desc(avg_inc))

```

```{r}
wealth_label %>%
  group_by(`Country Name`) %>%
  summarise(avg_cpc = 1000*mean(cpc)) %>%
  arrange(desc(avg_cpc))
```

